---
title: "intoASV: Workflow and QC (Heatmap-focused)"
author: "Heekuk Park"
date: "2026-01-23"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{intoASV workflow and QC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## 1. Overview

This vignette demonstrates a **standard intoASV workflow** with a focus on
**quality control (QC)** and **ASV-level distance heatmap visualization**.

The goal of this analysis is to quantify within-taxon microdiversity (π),
validate clustering and alignment quality, and generate both **static (PNG)**
and **interactive (HTML)** heatmaps for diagnostic purposes.

---

## 2. Running intoASV

The main intoASV analysis is performed using `Go_intoASV()`.
Below is an example configuration highlighting commonly used parameters.

```{r run-intoASV}
ps_intoASV <- Go_intoASV(
  psIN = ps2,
  project = project,

  # clustering and similarity thresholds
  clustering_cutoff = 0.995,
  similarity_cutoff = 0.97,

  # π calculation
  compute_ci = TRUE,
  method = "nucdiv",          # nucdiv: TN93-based approximation; simple: Hamming
  aligner = "DECIPHER",

  # filtering
  min_asv = 3,                # minimum ASVs per cluster
  min_abund = 10,             # minimum total abundance
  distance_gap = "exclude",

  # weighting scheme
  weighting = "entropy"       # entropy or abundance
)
```

---

## 3. QC 1: Cluster diagnostics

After clustering ASVs by sequence similarity, cluster-level diagnostics
should be examined to ensure that clusters represent meaningful biological
groupings rather than sequencing artifacts.

```{r qc-cluster}
Go_clusterDiagnostics(
  project = project,
  cluster_map =
    "intoASV_CP_260102/intoASV/pi_tab/cluster_map_similarity_0.970_260102.csv"
)
```

Recommended checks include:
- cluster size distribution
- abundance balance within clusters
- detection of singleton- or error-driven clusters

---

## 4. QC 2: Distance matrix and heatmap generation

To further inspect within-cluster heterogeneity, ASV-level distance matrices
can be visualized as heatmaps. These are particularly useful for identifying
outlier ASVs or residual redundancy after clustering.

### 4.1 Generate interactive heatmap (HTML)

The following **command-line workflow** generates an interactive HTML heatmap.
This file is intended for **exploratory analysis** and should be hosted
outside the R package (e.g. GitHub Pages).

```bash
conda activate qiime2

python3 ~/Dropbox/04_scripts/Go_intoASVs/intoASV/Python/asv_distance_heatmap.py \
  --dist  asv_distance_matrix_*.csv \
  --anno  cluster_diagnostics_*.csv \
  --order asv_order_*.csv \
  --out   ASV_distance_heatmap_$(date +%Y%m%d).html
```

The resulting HTML file provides:
- interactive zooming
- cluster-level annotations
- visual separation of closely related ASVs

---

### 4.2 Static heatmap for reproducibility

For reproducible documentation and package inclusion, a **static PNG version**
of the heatmap should be generated and stored under:

```
inst/figures/qc_asv_distance_heatmap.png
```

This static figure can be included directly in the vignette:

```{r qc-heatmap-static, echo=FALSE}
knitr::include_graphics(
  system.file("figures/qc_asv_distance_heatmap.png", package = "intoASV")
)
```

---

## 5. Interpretation notes

- Dense diagonal blocks indicate closely related ASVs within a cluster
- Off-diagonal structure may suggest sub-strain divergence
- Highly isolated ASVs should be re-evaluated for abundance or alignment quality

When combined with π estimates, these heatmaps provide a robust QC framework
to distinguish **intragenomic variation** from **true strain-level diversity**.

---

## 6. Summary

This vignette outlines a practical intoASV QC workflow, integrating:
- entropy-weighted π estimation
- cluster diagnostics
- ASV-level distance heatmaps (static and interactive)

Together, these steps ensure that inferred microdiversity reflects biological
signal rather than technical noise.
